---
title: "Gapminder"
author: "Joseph Vazhaeparampil"
date: 10-1-2026
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

- Country
- COntinent
- Year
- Life Expectancy
- Population
- GDP per capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(gapminder |> pull(year))
year_min <- min(gapminder |> pull(year))
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

q2 <- gapminder |>
  filter(year == year_min)

ggplot(q2, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 20000)) +
  labs(
    x = "Continent",
    y = "GDP per capita",
    title = paste("GDP per capita by continent in", year_min)
  )
```

**Observations**:

- Africa generally had the lowest GDP per capita, although there were definitely some variation
- Oceania seems to have the highest GDP per capita but was absed on a very samll number of datapoints, there is little variation
- Europe seems to have the largest amount of variation in GDP per capita (Standard deviation & range)

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them
  - The graph was initially not scaled well on the y-axis
  - We can observe the presence of several outliers, indicated by dots in the box plot

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
q2 |>
  group_by(continent) |>
  mutate(
    Q1 = quantile(gdpPercap, 0.25),
    Q3 = quantile(gdpPercap, 0.75),
    IQR = Q3 - Q1,
    outlier = gdpPercap < Q1 - 1.5 * IQR |
      gdpPercap > Q3 + 1.5 * IQR
  ) |>
  filter(outlier) |>
  select(country, continent, gdpPercap)
```

**Observations**:

- Identify the outlier countries from q2
  - We can see clearly that the outliers listed in the table are whats listed as outliers in the box plot

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
q4 <- gapminder |>
  filter(year %in% c(year_min, year_max))

ggplot(q4, aes(x = continent, y = gdpPercap)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 50000)) +
  geom_point(
    data = q4 |>
      group_by(year, continent) |>
      mutate(
        Q1 = quantile(gdpPercap, 0.25),
        Q3 = quantile(gdpPercap, 0.75),
        IQR = Q3 - Q1,
        outlier = gdpPercap < Q1 - 1.5 * IQR |
          gdpPercap > Q3 + 1.5 * IQR
      ) |>
      filter(outlier),
    aes(color = country),
    size = 2
  ) +
  facet_wrap(~year) +
  labs(
    x = "Continent",
    y = "GDP per capita",
    color = "Country"
  )
```

**Observations**:

- The United States is a consistent outlier in both 1952 and 2007 for GDP per cpaita in the Americas
- The range of GDP per capita has increased dramatically across all continents, but particularly the standard deviation in Africa and Asia

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder |>
  group_by(year, continent) |>
  summarize(avg_lifeExp = mean(lifeExp)) |>
  ggplot(aes(x = year, y = avg_lifeExp, color = continent)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Life Expectancy by Continent",
    x = "Year",
    y = "Average Life Expectancy",
    color = "Continent"
  )
```

- Average life expectancy seems to have increased across all of the continents
- Africa seems so have stopped increasing \~1987 and plateaued in terms of life expectancy
- It's slightly surprising to me how Oceania tops the list, but makes me wonder if we are not seeing the full picture with only 2 countries in Oceania. I wonder what it woudl look like if we had more datapoitns from the pacific island countries nearby Oceania.

```{r q5-task2}
## TASK: Your second graph

q5 <- gapminder |>
  filter(year %in% c(year_min, year_max))

ggplot(q5, aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    data = q5 |>
      group_by(year, continent) |>
      mutate(
        Q1 = quantile(lifeExp, 0.25),
        Q3 = quantile(lifeExp, 0.75),
        IQR = Q3 - Q1,
        outlier = lifeExp < Q1 - 1.5 * IQR |
          lifeExp > Q3 + 1.5 * IQR
      ) |>
      filter(outlier),
    aes(color = country),
    size = 2
  ) +
  facet_wrap(~year) +
  labs(
    x = "Continent",
    y = "Life Expectancy",
    color = "Country"
  )
```

- Afghanistan has a significantly lower life expectancy than the other countries in Asia in 2007 (perhaps due to political turmoil?)
- We observe dramatic increase in life expectancy in Asia and the Americas relative to the other continents
- Turkey seems to have recovered from a really low life expectancy relative to other Asian countries in the 50s. External research indicates that this may b e sue to lack of health infracture and the rampant spread of diseases in the post-WWII recovery era.

```{r q5-task3}
q5 <- gapminder |>
  filter(year %in% c(year_min, year_max))

ggplot(q5, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~year) +
  labs(
    title = "Life Expectancy vs. GDP per Capita",
    x = "GDP per capita (log scale)",
    y = "Life Expectancy",
    color = "Continent"
  )
```

- It seems pretty clear that higher GDP tends to be asociateed with higher life expectancy, especially more so in 2007 and in Europe, Asia, and the Americas
- It's surprising to me that several African countries with average GDPs per capita compared to the rest of the world still have some of the lowest life expectancy - makes me wonder about the welath distribution in these countries and who the GDP per capita actually benefits
- THis was my favorite graph as I think it reveleaded several interesting things that could be interesting to look further into
